### 2019-07-03 methodology/bayesian-networks.tex
Integrati TODO di AA per Cap. 3 e sezione su correlazione e IG

% !TEX root = thesis-thomas-tiotto.tex

\section{Mathematical background}
Bayesian Networks (BN) are a class of Probabilistic Graphical Models that are used to represent systems under conditions of uncertainty.
To give a formal definition we will first need a few basic concepts from probability and graph theory.

\subsection{Probability theory}
\subsubsection{Probability distributions}
\begin{definition}
	A \textit{probability distribution} is a function $\mathbb{P}: \mathcal{S} \rightarrow \mathbb{R}$ with $\mathcal{S}$ a set of \textit{events} of interest.  
To be a valid probability distribution $\mathbb{P}$ must satisfy:
\begin{itemize}
	\item $\mathbb{P}(\sigma) \geq 0 \quad \forall \sigma \in \mathcal{S}$
	\item $\sum_{\sigma} = 1 \quad \forall \sigma \in \mathcal{S}$
	\item $\alpha, \beta \in \mathcal{S} \wedge \alpha \cap \beta=\emptyset 	\Rightarrow \mathbb{P}(\alpha \cup \beta)=\mathbb{P}(\alpha)+\mathbb{P}(\beta)$
\end{itemize}
\end{definition}
Each event $\sigma \in \mathcal{S}$ must have a probability $\mathbb{P}(\sigma) \in [0,1]$ and the sum of all these must equal $1$. 
An event with $P(\sigma) = 0$ is deemed \textit{impossible} while one with $\mathbb{P}(\sigma) = 1$ is \textit{certain}.

@@

\subsubsection{Random variables}
\begin{definition}
	A random variable is a function that associates every outcome in $\mathcal{S}$ with a value.
\end{definition}
\textit{Random variables} are a way of bringing to the fore the attributes of interest of events while dealing with them in a clean, mathematical way.
The values that a random variable can take are a function of the events in sample space $\mathcal{S}$, each of these is assigned a value by the random variable function.
I will only be dealing with \textit{categorical random variables} i.e. those who's codomain is a discrete set of values.
Every random variable has a probability distribution induced by the cardinality of the subsets of its values; in the case of categorical-valued one, such a distribution is \textit{multinomial}.

\subsubsection{Conditional probabilities}
After having defined the basic notion of probability, we can construe one the basic building blocks of Bayesian Networks: the concept of \textit{conditional probability} 
\begin{definition}
	The conditional probability, ``the probability of event $\beta$ given event $\alpha$'' is:
\begin{equation} \label{eq:conditionalprobability}
\mathbb{P}(\beta \mid \alpha) = \frac{\mathbb{P}(\beta \cap \alpha)}{\mathbb{P}(\alpha)}
\end{equation}
\end{definition}
That is, the relative proportion of event $\beta$ compared to event $\alpha$; this intuitively represents the probability of $\beta$ \textit{knowing} that $\alpha$ has already occurred.

@@

\subsection{Graph theory}
Many problems in Machine Learning (ML) don't involve classification or prediction of single data points in isolation, but of set of entities that may present a more, or less, complex relation with each other. 
Most real-world phenomena fit into the latter framework.
Graphs are one of the most powerful tools for the modelling of this class of problems, as their structure naturally captures the wide variety of relations that may exist between entities.
These range from the atomical structure of a molecule to a social network of friends.  
In both these examples graphs help in reasoning, visualising and making inferences and predictions.

@@

\subsection{Bayesian Networks} \label{subsec:bayesiannetworks}
\begin{definition}
	A Bayesian Network (BN) is a probabilistic graphical model represented by a DAG where each vertex corresponds to a random variable $X_i$ and the edges model the dependencies among these.
\end{definition}
Such a model is basically a way of compactly representing an explicit joint distribution $\mathbb{P}(X_1 \cap \ldots \cap X_n) = \mathbb{P}(X_1) \ldots \mathbb{P}(X_n)$, that is factorised into $\mathbb{P}(X_n \mid X_1 \cap \ldots \cap X_{n-1}) \ldots \mathbb{P}(X_2 \mid X_1 ) \mathbb{P}(X_1) $.
The way this compactness is achieved is in exploiting the independencies that exist among the random variables:
\begin{equation} \label{eq:bnindependencies}
	\forall X_i:  ( X_i \perp \neg Desc(X_i) \mid Pa(X_i))
\end{equation}
with $Pa(X_i)$ the set of nodes that are parents of $X_i$ and $Desc(X_i)$ the nodes that are not \textit{descendents} of $X_i$.
That is to say, every random variable $X_i$, given its parent nodes, is independent of all other nodes in the Bayesian Network that are not descended from it.
Also, a BN gives the flexibility to drop the many weak dependencies that are bound to exist between variables thus leading to an even simpler model.
A full probability table for a joint distribution of random variables obscures the independencies and requires an exponential number of entries for the representation.
A Bayesian Network on the other hand can represent the same distribution using only a linear number of parameters.

@@

\subsubsection{d-separation}
The way that Bayesian Networks can be used to reduce the storage requirements for uncertain information is by taking advantage of the conditional independencies embedded in the underlying distribution being modelled.
The power of BNs comes from the additional information encoded in their structure and this was first explicitly described in its entirety by \cite{Pearl1988} who defined the concept of dependence separation, commonly known as \textit{d-separation}.

d-separation, as the name entails, talks about the conditional dependence between variables.
To define it, we first say when two variables $X$ and $Y$ are causally connected.
This is so if $Z = \emptyset$ and they are part of one of the following three structures:
\begin{itemize}
  \item $X \rightarrow Z \rightarrow Y$
  \item $X \leftarrow Z \leftarrow Y$
  \item $X \leftarrow Z \rightarrow Y$
\end{itemize}
This means that knowing something about $X$ also tells us something new about $Y$.
$X$ and $Y$ are causally independent if they appear in the following structure:
\begin{itemize}
  \item $X \rightarrow Z \leftarrow  Y$
\end{itemize}
Such a configuration is called a \textit{collider} and it blocks the flow of information from $X$ to $Y$.
If $Z \neq \emptyset$ the cases are reversed so colliders are open and the other three structures are blocked.
\begin{definition}
	Given disjoint subsets $X, Y, Z \subset \mathcal{X}$, $X$ and $Y$ are d-separated if:
	\begin{itemize}
		\item $Z \neq \emptyset$: no path between $X$ and $Y$ presents a collider
		\item $Z = \emptyset$: there is a collider on every path between $X$ and $Y$
	\end{itemize}
\end{definition}

\subsection{Bayesian Networks structure learning} \label{subsec:bnstructurelearning}
In many probabilistic models initialisation is fast but then fitting the data is slow (ex. k-means).
For Bayesian Networks the converse is true: fitting is fast as only sums of the counts in the data are needed but identifying the correct graph structure can take super-exponential time.
Learning the Bayesian Network structure from data is commonly known as the Bayesian Network Structure Learning (BNSL) problem.
The methods to solve this problem can be roughly categorised into one of three types.

@@

\subsubsection{Constraint learning}
Methods of this type calculate some measure of correlation to identify the presence and direction of edges between nodes.
A typical test is to iterate over all triplets while testing for conditional independencies.
Thanks to the d-separation properties outlined in Sub-Section. \ref{subsec:bayesiannetworks}, this test is able to identify the correct edges.
The algorithm is quadratic in time in the number of vertices.

@@

\subsection{Bayesian Networks updating} \label{subsec:bnupdating}
All the types of inference presented are instances of \textit{diagnostic reasoning}, also known as \textit{abductive reasoning}.  
This type of explanation can either be modelled as a conditional probability or a MAP query and is of fundamental importance in many important problems of machine learning including medical diagnosis, that is of particular interest to us.

\subsubsection{Conditional probability query}
The \textit{updating} problem is the process of updating the probabilities of nodes in the BN based on the observation of the values of other vertices.
This process of conditioning on observed information is also called \textit{data propagation}.

@@

\subsubsection{Maximum a posteriori query}
Another common type of question we might ask a BN is the following: ``given evidence $E$ which is the most likely assignment of a subset of variables $Y$?''.
This is know as \textit{Maximum a posteriori (MAP)} inference and is a much harder problem that a conditional probability query.
We are trying to solve the an optimisation problem.
\begin{definition}
As defined by \cite{koller2007introduction}.
Given evidence/observed variables $E=e$, $E \subseteq \mathcal{X}$ and sets $Y \subseteq \mathcal{X} - E$ and $Z = \mathcal{X} - E - Y$, with $\mathcal{X}$ the set of all variables in the BN, the MAP query for $Y$ is the assignment of values $Y=y$ that has maximum probability:
	\begin{equation} \label{eq:map}
	\text{MAP}( Y=y \mid E=e ) = \underset{y}{\text{argmax }}  \sum_z \mathbb{P}(Y=y, Z=z \mid E=e)
\end{equation}
\end{definition}

@@

\subsubsection{Most probable explanation query}
A special case of MAP is the \textit{Most probable explanation (MPE)} that, 
\begin{definition}
	As defined by \cite{koller2007introduction}.
 Given evidence/observed variables $E=e$, $E \subseteq \mathcal{X}$ and $W = \mathcal{X} - E$, the MPE query for $W$ is the assignment of values $W=w$ that has maximum probability:
	\begin{equation} \label{eq:mpe}
	\text{MPE}( W=w \mid E=e ) = \underset{w}{\text{argmax }} \mathbb{P}(W=w \mid E=e)
\end{equation}
\end{definition}

This is an easier problem than MAP, as can be seen by comparing Eq. \ref{eq:map} with Eq. \ref{eq:mpe}; MAP presents both a summation and a maximisation and as such is part conditional probability query, part MPE query.
All algorithms for the computation of MAP obviously apply to MPE too, but there exist efficient algorithms for MPE that do not generalise to MAP.
\todo{quali?}

\subsection{Information theory}
The birth of the field of \textit{information theory} is usually traced back to the seminal paper \enquote{A Mathematical Theory of Communication} (\cite{Shannon1949}) where Claude Shannon set the mathematical basis for the quantification of the amount of \textit{information} transmissible over a noisy channel. 
In his words \enquote{The fundamental problem of communication is that of reproducing at one point, either exactly or approximately, a message selected at another point.}
The concepts of field are broad enough to have influenced practically every other scientific discipline and deep enough to have enabled the \enquote{digital age}, for example by enabling the creation of ever more complicated coding schemes for the compression, reconstruction and obfuscation of digital data.

\subsubsection{Normalised entropy/efficiency}
Plain entropy is not a good choice when trying to characterise random variables with different cardinalities of their sample space.
Let us suppose that the objective is to find the variable with the least \enquote{entropic} distribution and we suppose that their values have all been generated by the same process, say Gaussian.
Simply calculating their entropies and ordering them according to this criterion will bias the selection process towards the variables with smallest cardinality.
This is because we supposed them to be distributed in the same way so there will naturally be less uncertainty when there are fewer possible outcomes.
This can easily be understood by imagining the distributions to all be random uniform.

To obviate to this problem we need to \textit{normalise} the entropy so that different-sized variables can be directly compared to each other.
To achieve this, we can look at a measure of normalised entropy or \textit{efficiency}:
\begin{equation} \label{eq:normalisedentropy}
 	\eta(X)=-\sum_{i=1}^{n} \frac{p\left(x_{i}\right) \log _{b}\left(p\left(x_{i}\right)\right)}{\log _{b}(n)}
\end{equation}
From Eq. \ref{eq:normalisedentropy} it can be seen that $\eta(X) \in [0,1]$; it is thus normalised and comparable among distributions.
This ratio expresses the amount of entropy found in the distribution compared to the maximum possible entropy when using $n$ symbols, corresponding to the uniform distribution:
\begin{equation}
\mathrm{H}\left(\underbrace{\frac{1}{n}, \ldots, \frac{1}{n}}_{n}\right) = - \sum_{i=1}^n \frac{1}{n} \log _{b} \left( \frac{1}{n} \right) = -n \cdot \frac{1}{n} \log _{b} \left( \frac{1}{n} \right) = - \log _{b} \left( \frac{1}{n} \right) = \log _{b}(n) 
\end{equation}

>>>>

% !TEX root = thesis-thomas-tiotto.tex
\todo{introduzione con outline del capitolo}
\todo{riassunto della fine del capitolo}

\section{Mathematical Background}
Bayesian Networks (BN) are a class of Probabilistic Graphical Models that are used to represent systems under conditions of uncertainty.
To give a formal definition we will first need a few basic concepts from probability and graph theory.

\subsection{Probability Theory}
\subsubsection{Probability distributions}
\begin{definition}
	A \textit{probability distribution} is a function $\mathbb{P}: \mathcal{S} \rightarrow \mathbb{R}$ with $\mathcal{S}$ a set of \textit{events} of interest.  
To be a valid probability distribution $\mathbb{P}$ must satisfy:
\begin{itemize}
	\item $\mathbb{P}(\sigma) \geq 0 \quad \forall \sigma \in \mathcal{S}$
	\item $\sum_{\sigma} = 1 \quad \forall \sigma \in \mathcal{S}$
	\item $\alpha, \beta \in \mathcal{S} \wedge \alpha \cap \beta=\emptyset 	\Rightarrow \mathbb{P}(\alpha \cup \beta)=\mathbb{P}(\alpha)+\mathbb{P}(\beta)$
\end{itemize}
\end{definition}
Each event $\sigma \in \mathcal{S}$ must have a probability $\mathbb{P}(\sigma) \in [0,1]$ and the sum of all these must equal $1$. 
An event with $P(\sigma) = 0$ is deemed \textit{impossible} while one with $\mathbb{P}(\sigma) = 1$ is \textit{certain}.

@@

\subsubsection{Random Variables}
\begin{definition}
	A random variable is a function that associates every outcome in $\mathcal{S}$ with a value.
\end{definition}
\textit{Random variables} are a way of bringing to the fore the attributes of interest of events while dealing with them in a clean, mathematical way.
The values that a random variable can take are a function of the events in sample space $\mathcal{S}$, each of these is assigned a value by the random variable function.
I will only be dealing with \textit{categorical random variables} i.e. those who's codomain is a discrete set of values.
Every random variable has a probability distribution induced by the cardinality of the subsets of its values; in the case of categorical-valued one, such a distribution is \textit{multinomial}.

If we were to take a Bayesian point of view, we would consider a random variable as simply representing the subjective degree of belief we would have over a set of outcomes we believed possible.

\subsubsection{Conditional Probabilities}
After having defined the basic notion of probability, we can construe one the basic building blocks of Bayesian Networks: the concept of \textit{conditional probability} 
\begin{definition}
	The conditional probability, ``the probability of event $\beta$ given event $\alpha$'' is:
\begin{equation} \label{eq:conditionalprobability}
\mathbb{P}(\beta \mid \alpha) = \frac{\mathbb{P}(\beta \cap \alpha)}{\mathbb{P}(\alpha)}
\end{equation}
\end{definition}
That is, the relative proportion of event $\beta$ compared to event $\alpha$; this intuitively represents the probability of $\beta$ \textit{knowing} that $\alpha$ has already occurred.

@@

\subsubsection{Correlation}
Correlation, as defined by \cite{Stolp2006}, is a measure of the degree to which two random variables are linearly dependent.
The most used measure of such a dependence is the \textit{Pearson Correlation Coefficient} or \textit{bivariate correlation}.
\begin{definition}
	The correlation coefficient $\rho$ of random variables $X$ and $Y$ is given by:
	\begin{align} \label{eq:correlation}
		\rho_{XY} &= \frac{Cov(X,Y)}{\sigma_X \sigma_Y} \\
		&= \frac{\mathbb{E}[ (X - \mu_X) (Y - \mu_Y) ]}{\sigma_X \sigma_Y} \\
		&= \frac{\sum_{x \in \mathcal{X}} \sum_{y \in \mathcal{Y}} (x - \mu_X) (y - \mu_Y) p_{XY}(x,y)}{ \sqrt{\sum_{x \in \mathcal{X}}  (x - \mu_X)^2 p_X(x)} \sqrt{\sum_{y \in \mathcal{Y}} (x - \mu_Y)^2 p_Y(y)} } 
	\end{align}
	with $p_{XY}$ the joint probability mass function of $X$ and $Y$ and $p_{X}$ and $p_{Y}$ the marginal distributions of $X$ and $Y$, respectively.
\end{definition}
That is, the bivariate correlation coefficient for random variables $X$ and $Y$ is given by the \textit{covariance} of $X$ and $Y$ divided by the product of their standard deviations.

The covariance is the \textit{first centred moment} of the \textit{joint distribution} of $X$ and $Y$ while the \textit{standard deviation} is the square root of the \textit{second centred moment} of the marginals.

$\rho_{XY}$ is normalised so its values vary in the interval $[-1,1]$; the correlation coefficient represents the degree of linear association between the two variables with $\rho_{XY}=-1$ being called \textit{perfect anticorrelation} and $\rho_{XY}=+1$ \textit{perfect correlation}.
The two correspond to the cases where the linear equation perfectly describes the relationship between $X$ and $Y$; the sign indicates the slope of the regression line describing the relationship i.e. if an increase in one of the two variables corresponds to an increase in the other in the pair, or viceversa.
The closer $\rho_{XY}$ tends to $0$, the feebler the relationship between $X$ and $Y$ with the case $\rho_{XY}=0$ indicating that the two variables are \textit{independent}.

\subsubsection{Mutual Information} \label{subsec:mutualinformation}
Another way of characterising the interrelatedness of two variables is through the concept of \textit{mutual information}, as defined in \cite{Cover2006}, that is closely linked to entropy, see Eq. \ref{eq:entropy}.
\begin{definition}
	The mutual information of two random variables $X$ and $Y$ is given by:
	\begin{equation}
		I_{XY} = \sum_{x \in \mathcal{X}} \sum_{y \in \mathcal{Y}} p_{XY}(x,y) \log \left( \frac{p_{XY}(x, y)}{p_{X}(x) p_{Y}(y)} \right)
	\end{equation}
\end{definition}
$I_{XY}$, intuitively, measures the amount of information that $X$ and $Y$ share that can also be seen as the degree to which one variable is informative of the other.
If $X$ and $Y$ are independent then they share no mutual information and knowing one of the two gives no information about the other.
if $X$ and $Y$ are perfectly correlated ($\rho_{XY}= \pm 1$) then they both convey the same amount of information and $I_{XY}$ is equal to the entropy $\Eta(X) = \Eta(Y)$.

\subsection{Information Theory}
The birth of the field of \textit{information theory} is usually traced back to the seminal paper \enquote{A Mathematical Theory of Communication} (\cite{Shannon1949}) where Claude Shannon set the mathematical basis for the quantification of the amount of \textit{information} transmissible over a noisy channel. 
In his words \enquote{The fundamental problem of communication is that of reproducing at one point, either exactly or approximately, a message selected at another point.}
The concepts of field are broad enough to have influenced practically every other scientific discipline and deep enough to have enabled the \enquote{digital age}, for example by enabling the creation of ever more complicated coding schemes for the compression, reconstruction and obfuscation of digital data.

\subsubsection{Normalised Entropy}
Plain entropy is not a good choice when trying to characterise random variables with different cardinalities of their sample space.
Let us suppose that the objective is to find the variable with the least \enquote{entropic} distribution and we suppose that their values have all been generated by the same process, say Gaussian.
Simply calculating their entropies and ordering them according to this criterion will bias the selection process towards the variables with smallest cardinality.
This is because we supposed them to be distributed in the same way so there will naturally be less uncertainty when there are fewer possible outcomes.
This can easily be understood by imagining the distributions to all be random uniform.

To obviate to this problem we need to \textit{normalise} the entropy so that different-sized variables can be directly compared to each other.
To achieve this, we can look at a measure of \textit{normalised entropy} or \textit{efficiency}:
\begin{equation} \label{eq:normalisedentropy}
 	\eta(X)=-\sum_{i=1}^{n} \frac{p\left(x_{i}\right) \log _{b}\left(p\left(x_{i}\right)\right)}{\log _{b}(n)}
\end{equation}
From Eq. \ref{eq:normalisedentropy} it can be seen that $\eta(X) \in [0,1]$; it is thus normalised and comparable among distributions.
This ratio expresses the amount of entropy found in the distribution compared to the maximum possible entropy when using $n$ symbols, corresponding to the uniform distribution:
\begin{equation}
\mathrm{H}\left(\underbrace{\frac{1}{n}, \ldots, \frac{1}{n}}_{n}\right) = - \sum_{i=1}^n \frac{1}{n} \log _{b} \left( \frac{1}{n} \right) = -n \cdot \frac{1}{n} \log _{b} \left( \frac{1}{n} \right) = - \log _{b} \left( \frac{1}{n} \right) = \log _{b}(n) 
\end{equation}

\subsection{Graph Theory}
Many problems in Machine Learning (ML) don't involve classification or prediction of single data points in isolation, but of set of entities that may present a more, or less, complex relation with each other. 
Most real-world phenomena fit into the latter framework.
Graphs are one of the most powerful tools for the modelling of this class of problems, as their structure naturally captures the wide variety of relations that may exist between entities.
These range from the atomical structure of a molecule to a social network of friends.  
In both these examples graphs help in reasoning, visualising and making inferences and predictions.

@@

\subsubsection{D-separation} \label{subsec:d-separation}
\textit{Dependence-separation} or \textit{d-separation}, as the name entails, is a concept relating to the conditional dependence between variables.
It was first presented by \cite{Pearl1988}
To define it, we first have to clarify when two sets of nodes $X$ and $Y$ are causally connected.
This is so if $Z = \emptyset$ and they are part of one of the following three structures, called \textit{v-structures} in this context:
\begin{itemize}
  \item $X \rightarrow Z \rightarrow Y$
  \item $X \leftarrow Z \leftarrow Y$
  \item $X \leftarrow Z \rightarrow Y$
\end{itemize}
This means that knowing something about $X$ also tells us something new about $Y$.
$X$ and $Y$ are causally independent if they appear in the following v-structure:
\begin{itemize}
  \item $X \rightarrow Z \leftarrow  Y$
\end{itemize}
Such a configuration is called a \textit{collider} and it blocks the flow of information from $X$ to $Y$.
If $Z \neq \emptyset$ the cases are reversed so colliders are open and the other three structures are blocked.
\begin{definition}
	Given disjoint subsets $X, Y, Z \subset \mathcal{X}$, $X$ and $Y$ are d-separated if:
	\begin{itemize}
		\item $Z \neq \emptyset$: no path between $X$ and $Y$ presents a collider
		\item $Z = \emptyset$: there is a collider on every path between $X$ and $Y$
	\end{itemize}
\end{definition}

An example using on the real network used in this thesis is shown in Fig. \ref{fig:d-separation_example}.
Here we can see how the network's topology and the nodes chosen to be in the observed set $Z$ define the resulting separations.
In this case $X= \{ \text{mut17q21} \}$, $Z= \{ \text{differenziazione, pN} \}$ and $Y=V \smallsetminus X \smallsetminus Z$; the resulting set of nodes that are d-separated from $X$ contains $\{ \text{c erB 2, FISH, loss 17, situ, pT, M, lateralita} \}$.
Basically, we are asking for the set of all nodes in the DAG that are d-separated from $X$, given evidence $Z$.
The reason for this can easily be given by enumerating all paths through the v-structures in the network and applying the definitions for causal connections given above:
\begin{itemize}
  \item mut17q21 $\rightarrow$ eta arrotondata: are directly connected so the cannot be independent
  \item mut17q21 $\rightarrow$ recettori estrogeni: like above
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ morfologia: recettori estrogeni is not in $Z$ so the path is open and mut17q21 is d-connected to morfologia
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ recettori progestinici: like above
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ ki67: as above
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ differenziazione: as above
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ differenziazione $\rightarrow$ c erbB 2: differenziazione is in $Z$ so this path is closed, as there are no other paths between mut17q21 and c erbB 2, the two nodes are d-separated
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ differenziazione $\rightarrow$ FISH: the only path to mut17q21 is through c erbB 2, that is already d-separated from mut17q21, so the two nodes are d-separated
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ differenziazione $\rightarrow$ FISH $\rightarrow$ loss 17: as above
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ differenziazione $\rightarrow$ pN: as recettori estrogeni, differenziazione and pN are arranged in a \textit{collider} and differenziazione is in $Z$, then the path is open
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ differenziazione $\rightarrow$ pN $\rightarrow$ situ: pN is in $Z$ and differenziazione, pN, situ are arranged as a \textit{fork}, thus the path is closed
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ differenziazione $\rightarrow$ pN $\rightarrow$ pT: as above, M is also d-separated from mut17q21 so there are no open paths between mut17q21 and pN
  \item mut17q21 $\rightarrow$ recettori estrogeni $\rightarrow$ differenziazione $\rightarrow$ pN $\rightarrow$ M: differenziazione, pN, M are arranged as a \textit{chain} and pN is in $Z$, the path is closed
  \item lateralita: lateralita is not connected to any other node and is thus d-separated from all others a priori
\end{itemize}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/d-separation-example}}
\caption{D-Separations in the provided data set (see Sec. \ref{sec:data-set}) generated using the \texttt{plot\_model} function (see Subsec. \ref{subsubsec:plot-model})}
\label{fig:d-separation_example}
\end{figure}

\subsection{Bayesian Networks} \label{subsec:bayesiannetworks}
\begin{definition}
	A Bayesian Network (BN) is a probabilistic graphical model represented by a DAG where each vertex corresponds to a random variable $X_i$ and the edges model the dependencies among these.
\end{definition}
Such a model is basically a way of compactly representing an explicit joint distribution $\mathbb{P}(X_1 \cap \ldots \cap X_n) = \mathbb{P}(X_1) \ldots \mathbb{P}(X_n)$, that is factorised into $\mathbb{P}(X_n \mid X_1 \cap \ldots \cap X_{n-1}) \ldots \mathbb{P}(X_2 \mid X_1 ) \mathbb{P}(X_1) $.
The way this compactness is achieved is in exploiting the independencies that exist among the random variables:
\begin{equation} \label{eq:bnindependencies}
	\forall X_i:  ( X_i \perp \neg Desc(X_i) \mid Pa(X_i))
\end{equation}
with $Pa(X_i)$ the set of nodes that are parents of $X_i$ and $Desc(X_i)$ the nodes that are not \textit{descendents} of $X_i$.
That is to say, every random variable $X_i$, given its parent nodes, is independent of all other nodes in the Bayesian Network that are not descended from it.
Also, a BN gives the flexibility to drop the many weak dependencies that are bound to exist between variables thus leading to an even simpler model.
A full probability table for a joint distribution of random variables obscures the independencies and requires an exponential number of entries for the representation.
A Bayesian Network on the other hand can represent the same distribution using only a linear number of parameters.
The way that Bayesian Networks can be used to reduce the storage requirements for uncertain information is by taking advantage of the conditional independencies embedded in the underlying distribution being modelled.
The power of BNs comes from the additional information encoded in their structure and this was first explicitly described in its entirety by \cite{Pearl1988} who defined the concept of dependence separation (see Subsec. \ref{subsec:d-separation}) and applied it to Bayesian Networks.

@@

\subsection{Bayesian Networks Structure Learning} \label{subsec:bnstructurelearning} 
In many probabilistic models initialisation is fast but then fitting the data is slow (ex. k-means).
For Bayesian Networks the converse is true: fitting is fast as only sums of the counts in the data are needed but identifying the correct graph structure can take super-exponential time.
Learning the Bayesian Network structure from data is commonly known as the Bayesian Network Structure Learning (BNSL) problem.
The methods to solve this problem can be roughly categorised into one of three types.

@@

\subsubsection{Constraint Learning}
Methods of this type calculate some measure of correlation to identify the presence and direction of edges between nodes.
A typical test is to iterate over all triplets while testing for conditional independencies.
Thanks to the d-separation properties outlined in Sub-Section. \ref{subsec:bayesiannetworks}, this test is able to identify the correct edges.
The algorithm is quadratic in time in the number of vertices.

@@

\subsection{Bayesian Networks Updating} \label{subsec:bnupdating}
All the types of inference presented are instances of \textit{diagnostic reasoning}, also known as \textit{abductive reasoning}.  
This type of explanation can either be modelled as a conditional probability or a MAP query and is of fundamental importance in many important problems of machine learning including medical diagnosis, that is of particular interest to us.

\subsubsection{Conditional Probability Query}
The \textit{updating} problem is the process of updating the probabilities of nodes in the BN based on the observation of the values of other vertices.
This process of conditioning on observed information is also called \textit{data propagation}.

@@

\subsubsection{Maximum a Posteriori Query}
Another common type of question we might ask a BN is the following: ``given evidence $E$ which is the most likely assignment of a subset of variables $Y$?''.
This is know as \textit{Maximum a posteriori (MAP)} inference and is a much harder problem that a conditional probability query.
We are trying to solve the an optimisation problem.
\begin{definition}
As defined by \cite{koller2007introduction}.
Given evidence/observed variables $E=e$, $E \subseteq \mathcal{X}$ and sets $Y \subseteq \mathcal{X} - E$ and $Z = \mathcal{X} - E - Y$, with $\mathcal{X}$ the set of all variables in the BN, the MAP query for $Y$ is the assignment of values $Y=y$ that has maximum probability:
	\begin{equation} \label{eq:map}
	\text{MAP}( Y=y \mid E=e ) = \underset{y}{\text{argmax }}  \sum_z \mathbb{P}(Y=y, Z=z \mid E=e)
\end{equation}
\end{definition}

@@

\subsubsection{Most Probable Explanation Query}
A special case of MAP is the \textit{Most probable explanation (MPE)} that, 
\begin{definition}
	As defined by \cite{koller2007introduction}.
 Given evidence/observed variables $E=e$, $E \subseteq \mathcal{X}$ and $W = \mathcal{X} - E$, the MPE query for $W$ is the assignment of values $W=w$ that has maximum probability:
	\begin{equation} \label{eq:mpe}
	\text{MPE}( W=w \mid E=e ) = \underset{w}{\text{argmax }} \mathbb{P}(W=w \mid E=e)
\end{equation}
\end{definition}

This is an easier problem than MAP, as can be seen by comparing Eq. \ref{eq:map} with Eq. \ref{eq:mpe}; MAP presents both a summation and a maximisation and as such is part conditional probability query, part MPE query.
All algorithms for the computation of MAP obviously apply to MPE too, but there exist efficient approximate algorithms for MPE that do not generalise to MAP such as Loopy Belief Propagation [\cite{Pearl1988}] and Stochastic Local Search [\cite{Kask1999}].

### 2019-07-03 methodology/data-set.tex
inizio sezione 3.3
Integrati TODO di AA per Cap. 3 e sezione su correlazione e IG

\section{Data set}
As anticipated in Chap. \ref{chap:introduction} the work carried out in this thesis had a certain degree of collaboration with a third party, \cite{istitutocantonalepresentazione}.

@@

\subsection{Provided data set}
The data set I was provided with consists of the histological records of 3218 breast cancer patients.
The data set had been pre-processed by collaborators of the ICT with some of the values being compiled by hand. 
In Tab. \ref{tab:datasetvariables} is a description of the measured variables, together with their clinical meaning.
The value distribution of the data set is shown in Tab. \ref{tab:datasetdistribution}.

@@

>>>>

\section{Data set} \label{sec:data-set}
As anticipated in Chap. \ref{chap:introduction} the work carried out in this thesis had a certain degree of collaboration with a third party, \cite{istitutocantonalepresentazione}.

@@

\subsection{Provided Data Set}
The data set I was provided with consists of the histological records of 3218 breast cancer patients.
The data set had been pre-processed by collaborators of the ICT with some of the values being compiled by hand. 
In Tab. \ref{tab:datasetvariables} is a description of the measured variables, together with their clinical meaning.
The value distribution of the data set is shown in Tab. \ref{tab:datasetdistribution}.

@@

\section{Validation Methodology}
\todo{discutere con Vittoria cosa vogliono esattamente e come validare }
